Clique and Constraint Models for Maximum Common (Connected) Subgraph Problems

### 2016-04-18 paper/paper.tex
Far too much mucking around (needs checking...)

% \usepackage{showframe}

\usepackage{microtype}                 % typographic perfection
\usepackage{complexity}                % \P, \NP etc
\usepackage{tikz}                      % For pretty pictures
\usepackage{amsmath}                   % \operatorname
\usepackage{amsfonts}                  % \mathbb
\usepackage{hyperref}                  % clicky links
\usepackage{cleveref}                  % no need to type Figure etc
\usepackage{gnuplot-lua-tikz}          % graphs
\usepackage[ruled,vlined]{algorithm2e} % algorithms (after cleverref!)
\usepackage{cite}                      % automatically sort mass cites
\usepackage[bottom]{footmisc}          % fix footnote + bottom figure on same page
\usepackage{placeins}

\usepackage{refcheck}

@@

\renewcommand{\floatpagefraction}{0.8}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.8}
\newcommand{\christine}[1]{{\color{green} Christine:#1}}

@@

Maximum common subgraph problems arise in biology and chemistry
\cite{DBLP:journals/jcamd/RaymondW02a,Ehrlich:2011}, in computer vision
\cite{DBLP:journals/jair/CookH94}, in the analysis of source code
\cite{DBLP:journals/tkde/DjokoCH97}, binary programs \cite{DBLP:conf/icics/GaoRS08}, and circuit
designs \cite{DBLP:journals/jair/CookH94}, in character recognition problems \cite{SIWEILU1991617},
and in many other domains \cite{Shasha:2002:AAT:543613.543620}, both directly and as a way of
measuring the similarity or difference between two graphs
\cite{DBLP:journals/prl/Bunke97,DBLP:journals/prl/FernandezV01,KriegeThesis}.

\paragraph{Definition of the Maximum Common (Connected) Subgraph Problem.}
We introduce definitions and algorithms on undirected and unlabelled graphs. The extension to
directed or labelled graphs is straightforward and discussed in Section \ref{extension}. An
undirected graph $G$ is defined by a finite set of nodes $N_G$ and a set of edges $E_G\subseteq
N_G\times N_G$ such that each edge is an undirected pair of nodes.

Given a graph $G$, two nodes $n_s, n_e \in N_G$ are connected by a path in $G$ if there exists a
sequence of nodes $<n_0,\ldots n_k>$ such that $n_0=n_s$, $n_k=n_e$ and each couple of successive
nodes is connected by an edge, {\em i.e.}, $\forall i\in [1,k], \{n_{i-1},n_i\}\in E_G$. A graph $G$
is \emph{connected} if every pair of nodes is connected by a path.

A graph $G$ is \emph{isomorphic} to another graph $G'$ if there exists a bijective function
$f:N_G\rightarrow N_{G'}$ which preserves edges, \emph{i.e.}, $\forall (u,v)\in N_G\times N_G,
\{u,v\}\in E_G\Leftrightarrow \{f(u),f(v)\}\in E_{G'}$.

A {\em subgraph} is obtained by removing nodes, and all their incident edges, {\em i.e.}, $G'$ is a
\emph{subgraph} of $G$ if $N_{G'}\subseteq N_G$ and $E_{G'}=E_G\cap(N_{G'}\times N_{G'})$.
In this definition, we only remove edges incident to removed nodes. This is usually called induced
subgraph in the litterature. We may also allow to remove edges which are not incident to removed
nodes, thus leading to partial subgraphs. In this paper, we only consider induced subgraphs.
However, all algorithms may be extended to the partial case, as explained in
\cite{DBLP:conf/cp/NdiayeS11}.

A \emph{common subgraph} of two graphs $G$ and $G'$ is a graph isomorphic to subgraphs of $G$ and
$G'$. A \emph{common connected subgraph} is a common subgraph which is connected.
A \emph{Maximum Common Subgraph} (MCS) (resp. \emph{Maximum Common Connected Subgraph} (MCCS) is a
common subgraph (resp. common connected subgraph) which has a maximum number of nodes. Examples of
MCS and MCCS are given in Fig. \ref{figure:mcsexample}.

\paragraph{Overview of the paper.} In Section \ref{existing}, we review existing approaches for
solving the MCS problem, with a specific focus on Constraint Programming (CP)-based approaches and
on clique-based approaches that reduce the problem to finding a maximum clique in an association
graph.

Conventional wisdom is that clique-based approaches are less effective in practice, but previous
experimental evaluations have used weak clique algorithms, or even maximal clique enumeration
algorithms.  Therefore, in Section \ref{eval1}, we  re-evaluate the clique-based approach using a
modern algorithm, and we show that it outperforms CP on labelled graphs, and that it is competitive
with CP on non labelled graphs.

Then, in Section \ref{mccs}, we consider the MCCS problem. For the CP approach, we may add a global
connectedness constraint to the model. Alternatively, we may use a special branching rule to grow
connected subgraphs only, as proposed in \cite{DBLP:conf/mco/VismaraV08}. These two techniques may
be combined, and we experimentally show that the best results are obtained when combining them.
When solving the MCCS problem with a clique-based approach, neither technique seems directly viable
with an association graph encoding.  However, we show that it is possible to adapt the combined
branching and bounding rule used by modern clique algorithms to preserve connectedness during
search.  We compare the clique-based approach with the best CP variant for the MCCS, and we show
that it outperforms CP on non labelled graphs, whereas it is outperformed by CP on labelled graphs.

@@

There exist two main approaches for solving MCS: The first approach (described in \ref{CP}) is based
on CP; The second approach (described in \ref{clique}) is based on a reformulation of MCS into a
maximum clique problem. Both approaches are described for non directed and non labelled graphs.
Their extension to directed or labelled graphs is discussed in Section \ref{extension}

\subsection{CP model for MCS}
\label{CP}

McGregor \cite{McGreg82} proposed a Branch \& Bound approach: Each branch of the search tree
corresponds to the matching of two nodes, and a  bounding function evaluates the number of nodes
that still may be matched so that the current branch is pruned as soon as this bound becomes lower
than the size of the largest known common subgraph. CP approaches may be viewed as enhancements of
this Branch \& Bound approach.

Vismara and Valery  \cite{DBLP:conf/mco/VismaraV08} introduced a first CP model. Given two graphs
$G$ and $G'$, this model associates a variable $x_u$ with every node $u$ of $G$, and the domain of
this variable contains all nodes of $G'$, plus an additional value $\bot$: variable $x_u$ is
assigned to $\bot$ if node $u$ is not matched to any node of $G'$; otherwise $x_u$ is assigned to
the node of $G'$ it is matched with. Edge constraints are introduced in order to ensure that
variable assignments preserve edges between matched nodes, {\em i.e.}, $\forall \{u,v\}\subseteq
N_G, (x_u=\bot)\vee (x_v=\bot) \vee (\{u,v\}\in E_G\Leftrightarrow \{x_u,x_v\}\in E_{G'})$.
Difference constraints are introduced in order to ensure that each node of $G'$ is assigned to at
most one variable, {\em i.e.}, $\forall \{u,v\}\subseteq N_G, (x_u=\bot)\vee (x_v=\bot) \vee
(x_u\neq x_v)$.

This CP model is improved in \cite{DBLP:conf/cp/NdiayeS11} by replacing binary difference
constraints with a soft global allDifferent constraint which maximizes the number of $x_u$ variables
that are assigned to values different from $\bot$, while ensuring they are all different when they
are not assigned to $\bot$.  Different constraint propagation techniques are compared in
\cite{DBLP:conf/cp/NdiayeS11}. The combination ``MAC+Bound'' (resp. ``FC+Bound'') obtains the best
results on labelled (resp. non labelled) graphs and outperforms the approach of \cite{McGreg82} and
the CP model of \cite{DBLP:conf/mco/VismaraV08}. The combination ``MAC+Bound'' maintains arc
consistency~\cite{sabi94} of edge constraints, whereas the conbination ``FC+Bound'' simply performs
forward-checking on these constraints. In both combinations, the ``Bound'' filtering checks whether
it is possible to assign distinct values to enough $x_u$ variables to surpass the best cost found so
far (it is a weaker version of GAC(\emph{softAllDiff}) \cite{peti01} which computes the maximum
number of variables that can be assigned distinct values).

\subsection{Reformulation of MCS as a maximum clique problem}
\label{clique}

We may solve MCS by introducing an association graph and searching for a maximum clique in it
\cite{bala86,dura99,DBLP:journals/jcamd/RaymondW02a}.

An association graph of two graphs $G$ and $G'$ is an undirected graph $A_{G,G'}$ whose set of nodes
is $N_{A_{G,G'}} = N_G \times N_{G'}$. To avoid confusing nodes of $A_{G,G'}$ with initial graph
nodes, nodes of $A_{G,G'}$ are called matching nodes, as each node $(u,u')$ of $A_{G,G'}$ denotes
the matching of $u$ with $u'$. Edges of $A_{G,G'}$ connect matching nodes which are compatible, {\em
i.e.}, $E_{A_{G,G'}} = \big\{\{(u,u'),(v,v')\}\subseteq N_{A_{G,G'}}$, $(u,u')$ and $(v,v')$ are
compatible$\big\}$, where two matching nodes $(u,u')$ and $(v,v')$ of $N_{A_{G,G'}}$ are compatible
if $u \neq v$ and $u' \neq v'$, and if they preserve edges (\emph{i.e.}, $\{u,v\} \in E_G
\Leftrightarrow \{u',v'\} \in E_{G'}$).

A \emph{clique} is a subgraph whose nodes are all linked pairwise.  A clique is \emph{maximal} if it
is not strictly included in any other clique, and it is \emph{maximum} if it is the largest clique
of a given graph, with respect to the number of nodes.

A clique in $A_{G,G'}$ corresponds to a set of compatible matchings. Therefore, such a clique
corresponds to a common subgraph, and a maximum clique of $A_{G,G'}$ is a MCS of $G$ and $G'$.  It
follows that any method able to find a maximum clique in a graph can be used to solve the MCS
problem.

Note that the association graph is a partial subgraph of the microstructure
\cite{DBLP:conf/aaai/Jegou93a} associated with the CP model of  \cite{DBLP:conf/mco/VismaraV08}: The
microstructure has more matching nodes than the association graph because it has a matching node
$(u,\bot)$ for each node $u$ of $G$. Each clique of size $|N_G|$ in the microstructure corresponds
to a common subgraph the size of which is defined by the number of nodes that do not contain $\bot$.

\subsection{Extension to Labeled or Directed Graphs}\label{extension}

In some applications, nodes and edges are associated with labels. We note $\lambda(u)$ and
$\lambda(\{u,v\})$ the label of a node $u$ and an edge $\{u,v\}$, respectively. In this case, the
common  subgraph must match components the labels of which are compatible, and we note $\simeq$ the
compatibility relationship, {\em i.e.}, labels $\lambda_1$ and $\lambda_2$ are compatible if
$\lambda_1\simeq\lambda_2$. This kind of label compatibility constraint is handled in a
straightforward way in both CP and clique-based approaches:

\begin{itemize}
\item For CP, we restrict the domain of every variable $x_u$ to nodes with compatible labels, {\em
    i.e.}, $\forall u\in N_G, D(x_u) = \{\bot\}\cup \{u'\in N_{G'}, \lambda(u)\simeq\lambda(u')\}$.
    We also have to ensure that edge labels are preserved in edge constraints, {\em i.e.}, $\forall
    \{u,v\}\subseteq N_G, (x_u=\bot)\vee (x_v=\bot) \vee ((\{u,v\}\in E_G\Leftrightarrow
    \{x_u,x_v\}\in E_{G'}) \wedge (\lambda(\{u,v\}) \simeq \lambda(\{x_u,x_v\})))$.

\item For clique-based approaches, label compatibility is handled through the definition of the
    association graph, by restricting the set of matching nodes to pairs of nodes with compatible
    labels, {\em i.e.}, $N_{G_C} = \{(u,u')\in N_G\times N_{G'}, \lambda(u)\simeq\lambda(u')\}$.
    Also, edges of the association graph are restricted to pairs $\{(u,u'),(v,v')\}$ such that
    $\lambda(\{u,v\}) \simeq \lambda(\{u',v')\}$.
\end{itemize}

Labels usually simplify the solution process, both for CP and clique-based approaches: Node labels
reduce domain sizes for CP, and the number of matching nodes in the association graphs; Edge labels
tighten edge constraints for CP, and make the association graph sparser for clique-based approaches.

The extension of MCS algorithms to directed graphs is straightforward: For CP, edge orientation is
handled in edge constraints; for clique-based approaches, it is handled in the definition of the
association graph edges. In both cases, adding orientations to edges usually simplifies the problem:
It tightens edge constraints for CP, and it makes the association graph sparser for  clique-based
approaches.

@@

Previous experimental work has used a maximal clique enumeration algorithm, even for the
maximisation problem. ?? check this! I recall someone using a bad max clique algorithm too.
\cite{DBLP:conf/sspr/BunkeFGSV02,DBLP:journals/jgaa/ConteFV07}

We instead use a maximum clique algorithm.

Also sparse stuff, but association graphs aren't sparse enough.

Say which clique algorithm we're using.

\subsection{Datasets}

We consider a randomly generated database described in
\cite{DBLP:journals/prl/SantoFSV03,DBLP:journals/jgaa/ConteFV07}.
The dataset contains different classes of graphs: Randomly connected graphs with connectivity
$\eta\in\{0.05,0.2\}$ (r005 and r02); 2D, 3D, and 4D regular meshes (m2D, m3D, m4D); 2D, 3D, and 4D
irregular meshes with $\rho=0.6$ (m2Dr, m3Dr, and m4Dr); regular bounded valence graphs with
$V\in\{3,9\}$ (b03 and b09) and irregular bounded valence graphs with $V\in\{3,9\}$ (b03m and b09m).
Each class contains XXX pairs of graphs: YYY for each of the 5 possible sizes of the MCIS ({\em
i.e.}, 10\%, 30\%, 50\%, 70\% and 90\% of the number of nodes of the original graphs).

For each pair of graphs, there are 3 different labelings such that the number of different labels is
equal to 33\%, 50\% or 75\% of the number of nodes. In this paper, we report experiments with non
labelled graphs (labels are ignored), and with 33\% labellings.

\subsection{Experimental Evaluation}

@@

    \centering
    \input{gen-graph-unconnected-heatmap}
    \caption{The cumulative number of instances solved in under a certain time: on the top, 33\%
        labelled graphs with up to 100 vertices, and in the middle, unlabelled and undirected graphs
        with up to 35 vertices. On the bottom row, an instance-by-instance comparison of the clique
        model with the best CP model, with 33\% labelled graphs (with MAC) on the left, and
        unlabelled and undirected graphs (with FC) on the right.} \label{figure:unconnected-cumulative}
\end{figure}

Fig. \ref{figure:unconnected-cumulative} displays the cumulative number of instances solved with respect to time.
When graphs are labelled (top plot), the clique-based approach clearly outperforms CP. ...develop...

\subsection{Parallel Search}

\cite{DBLP:journals/jcc/KoncDTRJ12,DBLP:journals/algorithms/McCreeshP13,DBLP:journals/topc/McCreeshP15,DBLP:journals/cor/SegundoLP16}

?? Stuff on EHPs, how work stealing matters here too \cite{DBLP:journals/jco/BatsynGMP14}

@@

In many applications, the common subgraph must be connected. As illustrated in Fig.
\ref{figure:mcsexample}, the MCCS cannot be deduced from the MCS: We need to ensure connectedness
during the search. In Section \ref{mccs-cp}, we show that this may be done in two different ways
that may be combined, and we show in Section \ref{mccs-cp-eval} that the best results are obtained
when combining them. In Section \ref{mccs-clique}, we introduce a new way for ensuring connectedness
in a clique-based approach. Finally, we compare CP and our clique-based approach in Section
\ref{mccs-eval}.

@@

\paragraph{Ensuring connectedness with a specific branching rule.} In
\cite{DBLP:conf/mco/VismaraV08}, the connectedness constraint is ensured by using a specific
branching rule, used to select the next variable to be assigned. Let $CComp$ be the set of variables
already assigned to values different from $\bot$. The next variable to be assigned is chosen within
the set of non assigned variables which are neighbor of at least one vertex of $CComp$. When this
set is empty, all remaining non assigned variables are assigned to $\bot$.

\paragraph{Ensuring connectedness with a specific filtering.}

CP(Graph) \cite{DBLP:conf/cp/DoomsDD05} introduces graph domain variables and enforces connectivity via the
\emph{reachable} constraint, ensuring that there is a path from a specified vertex to a specified
set of vertices. One such constraint is then posted for each of the vertices in the graph and
encodes the transitive closure of the graph. Brown et al.\ \cite{Brown:2005} explored the use of
constraint programming in the generation of connected graphs with specified degree sequences.  Two
constraints were combined, i.e.\ the graphical constraint (a backtrackable implementation of the
Havel-Hakimi algorithm) with a connectivity constraint implemented using sets of vertices, where
vertex sets $A$ and $B$ are combined when there exists a pair of vertices $v \in A$ and $w \in B$
and an edge $(v,w) \in E$. Residual degree counts are maintained on components and vertices to
enforce graphicality and connectivity. In \cite{DBLP:conf/ecai/ProsserU06} a connectivity constraint was proposed for
connected graph generation where decision variables are edges (the search process accepts and
rejects edges). The constraint employed depth first search to maintain the set of tree edges and
back edges, associating path counters on these edges. The counters were then used to detect the
existence of cut-edges and protects these by forcing edges. The complexity of this constraint is
cubic in the number of vertices.

In all these previous works, the goal is to ensure that a given set of vertices is connected. For
MCCS, the problem is slightly different: we have to ensure that the number of connected vertices
that may be matched (in both graphs) is greater than the size of the largest common subgraph
previously found. Therefore, we introduce a new filtering algorithm to ensure connectedness
consistency for MCCS.  Let us consider two graphs $G$ and $G'$, and let $D$ be the current domains
(we suppose that $D(x_u)$ is a singleton when $x_u$ is assigned). Let $C$ (resp. $C'$) be the set of
nodes of $G$ (resp. $G'$) which may belong to the common subgraph, {\em i.e.}, $C=\{ u\in N_{G},
D(x_u)\neq\{\bot\}\}$ (resp. $C' = \{u'\in N_{G'}, \exists u\in N_G, u'\in D(x_u)\}$). Let
$G_{\downarrow C}$ (resp. $G'_{\downarrow C'}$) be the subgraph of $G$ induced by $C$ (resp. the
subgraph of $G'$ induced by $C'$).  The connectedness consistency ensures that both $G_{\downarrow
C}$ and $G'_{\downarrow C'}$ are connected graphs.

Connectedness consistency is not ensured at the root node of the search. It is ensured only once a
first variable has been assigned. Let $x_u$ be the first assigned variable and $u'$ the value
assigned to $x_u$. To ensure connectedness consistency, we perform a traversal of $G$ (resp. $G'$),
starting from $u$ (resp. $u'$), and we initialize $C$ (resp. $C'$) with all visited nodes. Then, for
each node $v\in N_G\setminus C$ (resp. $v'\in N_{G'}\setminus C'$), we assign $x_v$ to $\bot$
(resp., we remove $v'$ from all domains it belongs to).

During the search, each time a variable is assigned to $\bot$, we remove the corresponding node from
$C$ and perform a new traversal of $G_{\downarrow C}$ starting from the initial node $u$. For each
node $v\in C$ that is not visited by the traversal, we remove $v$ from $C$ and assign $x_v$ to
$\bot$. Also, each time a value is removed from a domain so that this value no longer belongs to any
domain, we remove it from $C'$, and perform a new traversal of $G'_{\downarrow C'}$ starting from
the initial node $u'$. For each node $v'$ that is not visited by the traversal, we remove $v'$ from
$C'$, and remove $v'$ from all domains it belong to.

\paragraph{Combining both approaches.} The two approaches for ensuring connectedness (branching and
filtering) are complementary and may be combined: At each step of the search, we select the next
variable to be assigned within the neighbors of $CComp$, and each time a node of $G'$ is removed
from a domain we filter domains to ensure connectedness consistency.

>>>>

% \usepackage{showframe}
% \usepackage{refcheck}

\usepackage{microtype}                 % typographic perfection
\usepackage{complexity}                % \P, \NP etc
\usepackage{tikz}                      % For pretty pictures
\usepackage{amsmath}                   % \operatorname
\usepackage{amssymb}                   % \triangledown
\usepackage{hyperref}                  % clicky links
\usepackage{cleveref}                  % no need to type Figure etc
\usepackage{gnuplot-lua-tikz}          % graphs
\usepackage[ruled,vlined]{algorithm2e} % algorithms (after cleverref!)
\usepackage{cite}                      % automatically sort mass cites
\usepackage[bottom]{footmisc}          % fix footnote + bottom figure on same page

@@

\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{0.85}
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.8}

\newcommand{\christine}[1]{{\color{green} Christine:#1}}

\newcommand{\modularproduct}{\operatorname{\triangledown}}

@@

Maximum common subgraph problems arise in biology and chemistry
\cite{DBLP:journals/jcamd/RaymondW02a,Ehrlich:2011}, in computer vision
\cite{DBLP:journals/jair/CookH94}, in the analysis of source code
\cite{DBLP:journals/tkde/DjokoCH97}, binary programs \cite{DBLP:conf/icics/GaoRS08}, and circuit
designs \cite{DBLP:journals/jair/CookH94}, in character recognition problems \cite{SIWEILU1991617},
and in many other domains \cite{Shasha:2002:AAT:543613.543620}, both directly and as a way of
measuring the similarity or difference between two graphs
\cite{DBLP:journals/prl/Bunke97,DBLP:journals/prl/FernandezV01,KriegeThesis}. We illustrate two
variants of this problem in \cref{figure:mcsexample}.

\subsection{Definitions and Notation}

We introduce definitions and algorithms on undirected and unlabelled graphs without loops; the
extension to general graphs is straightforward and is discussed in \cref{extension}. An undirected
graph $G$ is defined by a finite set of vertices $\operatorname{V}(G)$ and a set of undirected edges
$\operatorname{E}(G) \subseteq \operatorname{V}(G) \times \operatorname{V}(G)$, where $(u, v) \in
\operatorname{E}(G) \Rightarrow (v, u) \in \operatorname{E}(G) \wedge u \ne v$. The neighbourhood of
a vertex $v$, written $\operatorname{N}(G, v)$, is the set of vertices to which it is adjacent.
Given a graph $G$, two vertices $v_s, v_e \in \operatorname{V}(G)$ are connected by a path in $G$ if
there exists a sequence of vertices $(v_0, v_1, \ldots, v_k)$ such that $v_0 = v_s$, $v_k = v_e$,
and each pair of successive vertices is connected by an edge, i.e.\ $\forall i\in \{1 \ldots k \},
\{v_{i-1},v_i\}\in \operatorname{E}(G)$. A graph $G$ is \emph{connected} if every pair of vertices
is connected by a path.

The \emph{subgraph} of a graph $G$ \emph{induced by} a set $H \subseteq \operatorname{V}(G)$,
written $G[H]$, is the graph with vertex set $H$, and with every edge in $G$ which has both
endpoints in $G$, i.e.\ $\operatorname{E}(H) = \operatorname{E}(G) \cap (\operatorname{V}(H) \times
\operatorname{V}(H))$. We will consider only subgraphs which are induced by some set. It is also
possible to permit removing edges which are not incident to removed vertices, thus leading to
partial subgraphs---we do not consider this possibility in this paper, although everything we
discuss may be extended to the partial case \cite{DBLP:conf/cp/NdiayeS11}.

A graph $G$ is \emph{isomorphic} to another graph $H$ if there exists a bijective function $f :
\operatorname{V}(G) \rightarrow \operatorname{V}(H)$ which preserves edges and non-edges, i.e.\
$\forall (u, v) \in \operatorname{V}(G) \times \operatorname{V}(G), (u, v) \in \operatorname{E}(G)
\Leftrightarrow (f(u),f(v)) \in \operatorname{E}(H)$.  A \emph{common subgraph} of two graphs $G$
and $H$ is a graph isomorphic to subgraphs of $G$ and $H$. A \emph{common connected subgraph} is a
common subgraph which is connected.  A \emph{Maximum Common Subgraph} (MCS) (resp. \emph{Maximum
Common Connected Subgraph} (MCCS) is a common subgraph (resp.\ common connected subgraph) which has
a maximum number of vertices.

\subsection{Overview}

In \cref{existing}, we review existing approaches for solving the MCS problem, with a specific focus
on Constraint Programming (CP)-based approaches, and on approaches that reduce the problem to
finding a maximum clique in an association graph.

Conventional wisdom is that clique-based approaches are less effective in practice, but previous
experimental evaluations have used weak clique algorithms, or even maximal clique enumeration
algorithms.  Therefore, in \cref{eval1}, we  re-evaluate the clique-based approach using a modern
algorithm, and we show that it outperforms CP on labelled graphs, and that it is competitive with CP
on unlabelled graphs.

Then, in \cref{mccs}, we consider the MCCS problem. For the CP approach, we may add a global
connectedness constraint to the model. Alternatively, we may use a special branching rule
\cite{DBLP:conf/mco/VismaraV08} to grow connected subgraphs only. These two techniques may
be combined, and we experimentally show that the best results are obtained when combining them.
When solving the MCCS problem with a clique-based approach, neither technique seems directly viable
with an association graph encoding.  However, we show that it is possible to adapt the combined
branching and bounding rule used by modern clique algorithms to maintain connectedness during
search.  We compare the clique-based approach with the best CP variant for the MCCS, and we show
that it outperforms CP on labelled graphs, whereas it is outperformed by CP on unlabelled graphs.

@@

There are two main approaches for solving MCS. The first approach (described in \cref{CP}) is based
on CP, whilst the second (described in \cref{clique}) is based on a reformulation of MCS into a
maximum clique problem. Both approaches are described for non directed and non labelled graphs;
their extension to directed or labelled graphs is discussed in \cref{extension}.

\subsection{Constraint Programming Models for MCS} \label{CP}

McGregor \cite{McGreg82} proposed a branch and bound approach: each branch of the search tree
corresponds to the matching of two vertices, and a bounding function evaluates the number of
vertices that still may be matched so that the current branch is pruned as soon as this bound
becomes lower than the size of the largest known common subgraph. CP approaches may be viewed as
enhancements of this branch and bound algorithm.

Vismara and Valery  \cite{DBLP:conf/mco/VismaraV08} introduced the first explicit CP model. Given
two graphs $G$ and $H$, this model associates a variable $x_v$ with every vertex $v$ of $G$, and the
domain of this variable contains all vertices of $H$, plus an additional value $\bot$: variable
$x_v$ is assigned to $\bot$ if vertex $v$ is not matched to any vertex of $H$; otherwise $x_v$ is
assigned to the vertex of $H$ to which it is matched. Edge constraints are introduced in order to
ensure that variable assignments preserve edges and non-edges between matched vertices, i.e.\ $\forall u, v \in
\operatorname{V}(G), (x_u = \bot) \vee (x_v = \bot) \vee ((u, v) \in \operatorname{E}(G)
\Leftrightarrow (x_u, x_v) \in \operatorname{E}(H))$.  Difference constraints are introduced in
order to ensure that each vertex of $H$ is assigned to at most one variable, i.e.\ $\forall u, v \in
\operatorname{V}(G)~\textnormal{distinct,}~ (x_u = \bot) \vee (x_v = \bot) \vee (x_u \neq x_v)$.

This CP model is improved by Ndiaye and Solnon \cite{DBLP:conf/cp/NdiayeS11} by replacing binary
difference constraints with a soft global allDifferent constraint which maximizes the number of
$x_u$ variables that are assigned to values different from $\bot$, while ensuring they are all
different when they are not assigned to $\bot$. They find that the combination ``MAC+Bound'' (resp.
``FC+Bound'') obtains the best results on labelled (resp. non labelled) graphs and outperforms the
two previous approaches. The combination ``MAC+Bound'' maintains arc consistency~\cite{sabi94} of
edge constraints, whereas the combination ``FC+Bound'' simply performs forward-checking on these
constraints. In both combinations, the ``Bound'' filtering checks whether it is possible to assign
distinct values to enough $x_u$ variables to surpass the best cost found so far (it is a weaker
version of GAC(\emph{softAllDiff}) \cite{peti01} which computes the maximum number of variables that
can be assigned distinct values).

\subsection{Reformulation of MCS to a Maximum Clique Problem}\label{clique}

An alternative approach to MCS is to reduce the problem to finding a maximum clique in an
association graph \cite{LeviG,bala86,dura99,DBLP:journals/jcamd/RaymondW02a}.
An association graph (or compatibility graph, or weak modular product) of two graphs $G$ and $H$ is
an undirected graph $G \modularproduct H$ with vertex set $\operatorname{V}(G \modularproduct H) =
\operatorname{V}(G) \times \operatorname{V}(G)$---to avoid confusing vertices of $G \modularproduct
H$ with vertices of the two original graphs, we call vertices of $G \modularproduct H$
\emph{matching nodes}, as each vertex $(u, v)$ of $G \modularproduct H$ denotes the matching of $u$
with $v$. The edges of $G \modularproduct H$ connect matching nodes which denote compatible
assignments, so two matching nodes $(u, u')$ and  $(v, v')$ are adjacent if $u \neq v$ and $u' \neq
v'$, and if they preserve both edges and non-edges, so $(u, v) \in \operatorname{E}(G) \Leftrightarrow
(u', v') \in \operatorname{E}(H)$.

A \emph{clique} is a subgraph whose vertices are all pairwise adjacent. A clique is \emph{maximal}
if it is not strictly included in any other clique, and it is \emph{maximum} if it is a largest
clique of a given graph, with respect to the number of vertices.  A clique in an association graph
corresponds to a set of compatible matchings. Therefore, such a clique corresponds to a common
subgraph, and a maximum clique of $G \modularproduct H$ is an MCS of $G$ and $H$.  It follows that
any method able to find a maximum clique in a graph can be used to solve the MCS problem.

Note that the association graph is a partial subgraph of the microstructure
\cite{DBLP:conf/aaai/Jegou93a} associated with the CP model of  Vismara and Valery
\cite{DBLP:conf/mco/VismaraV08}: the microstructure has more matching nodes than the association
graph because it has a matching node $(u,\bot)$ for each vertex $u$ of $G$. Each clique of size
$\left|\operatorname{V}(G)\right|$ in the microstructure corresponds to a common subgraph, the size
of which is defined by the number of matching nodes that do not contain $\bot$.

\subsection{Extension to Labelled or Directed Graphs, and Loops}\label{extension}

In some applications, labels may be associated with vertices or edges. We denote $\lambda(u)$ and
$\lambda((u, v))$ the label of a vertex $u$ and an edge $(u, v)$, respectively. Where graphs are
labelled, any isomorphism $f$ must additionally preserve labels, so we require $\lambda(f(v)) =
\lambda(v)$ for any vertex $v$, and $\lambda((f(u), f(v))) = \lambda((u, v))$ for any edge $(u, v)$.
This kind of label compatibility constraint is handled in a straightforward way in both CP and
clique-based approaches to MCS.  For CP, we restrict the domain of every variable $x_u$ to vertices
with compatible labels, and ensure that edge labels are preserved in edge constraints.
For clique-based approaches, label compatibility is handled through the definition of the
association graph, by restricting the set of matching nodes to pairs of vertices with compatible
labels, and the set of matching edges to pairs of edges with compatible labels.

The extension of MCS algorithms to directed graphs, where isomorphisms must preserve directed edges,
is similarly straightforward.
Finally, some datasets contain loops (self-adjacent vertices). In this case vertices with loops must
be mapped to vertices with loops (with the same labels, if appropriate), and vertices without loops
may not be mapped to vertices with loops.

Labels, directed edges and loops usually simplify the solution process, both for CP and clique-based
approaches: vertex labels reduce domain sizes for CP, and the number of matching nodes in the
association graphs; edge labels tighten edge constraints for CP, and make the association graph
sparser for clique-based approaches.

?? More on AC here? If we think carefully about the size of the domains of the objective variable:
tightening the constraints doesn't really help unless you've made an assignment, right?

@@

Previous experimental evaluations of the association graph model have used either maximal clique
enumeration algorithms \cite{DBLP:journals/tcs/Koch01,DBLP:conf/mco/VismaraV08} (even when the
maximisation problem was being considered), or very simple maximum clique algorithms
\cite{DBLP:conf/sspr/BunkeFGSV02,DBLP:journals/jgaa/ConteFV07}. Thus we re-evaluate the approach
using a modern maximum clique algorithm. Association graphs are dense, even if the input is sparse,
so we will using (the single-threaded, bit-parallel version of) the maximum clique solver by
McCreesh and Prosser \cite{DBLP:journals/topc/McCreeshP15}, which implements Prosser's
\cite{DBLP:journals/algorithms/Prosser12} ``MCSa1'' variant of a series of algorithms due to Tomita
\textit{et al.}\
\cite{DBLP:conf/dmtcs/TomitaS03,DBLP:journals/jgo/TomitaK07,DBLP:conf/walcom/TomitaSHTW10}, using a
bitset encoding due to San Segundo \textit{et al.}\
\cite{DBLP:journals/cor/SegundoRJ11,DBLP:journals/ol/SegundoMRH13}. We compare this to the CP
implementation of Ndiaye and Solnon \cite{DBLP:conf/cp/NdiayeS11}.  We perform our experiments on
machines with Intel Xeon E5-2640 v2 CPUs and 64GBytes RAM; software was compiled using GCC 4.9, and
a timeout of one hour was used.

We consider a randomly generated database
\cite{DBLP:journals/prl/SantoFSV03,DBLP:journals/jgaa/ConteFV07} commonly used for benchmarking
maximum common subgraph problems.  The dataset contains different classes of graphs: randomly
connected graphs with different densities; 2D, 3D, and 4D regular and irregular meshes; regular
bounded valence graphs, and irregular bounded valence graphs.  For each pair of graphs, there are 3
different labellings such that the number of different labels is equal to 33\%, 50\% or 75\% of the
number of vertices. In this paper, we report experiments with unlabelled graphs (labels are
ignored), and with 33\% labellings. For unlabelled graphs, we select the 27,500 graph pairs where the
number of vertices in each graph is no more than 35; for labelled graphs, which we find less
computationally challenging, we select all 81,400 graph pairs, to include graphs with up to 100
vertices.

@@

    \centering
    \input{gen-graph-unconnected-heatmap}
    \caption{The cumulative number of instances solved in under a certain time: on the top, 33\%
        labelled graphs, and in the middle, unlabelled and undirected graphs. On the bottom row, an
        instance-by-instance comparison of the clique model with the best CP model, with 33\%
        labelled graphs (with MAC) on the left, and unlabelled and undirected graphs (with FC) on
    the right.} \label{figure:unconnected-cumulative}
\end{figure}

The top two plots of \cref{figure:unconnected-cumulative} display the cumulative number of instances
solved with respect to time.  When graphs are labelled, the clique-based approach clearly
outperforms either CP model, and MAC has a slight advantage over FC. For unlabelled graphs, the
three approaches are broadly comparable, and ultimately FC beats MAC, which beats the clique
approach. The bottom row gives a per-instance comparison of the best CP approach with the clique
approach: the heatmaps are similar to scatter plots, but due to the large number of instances, we
colour each point according to the density of solutions around that point. For labelled graphs, the
clique approach comes close to dominating MAC on non-trivial instances (which suggests that there is
unlikely to be scope for per-instance algorithm selection here). For unlabelled graphs, there is
still a broad correlation between the runtimes; the clique approach rarely wins by more than one
order of magnitude, but is sometimes much worse.

A closer inspection of the data suggests that the different randomness models have little effect on
the runtimes for either approach. However, the relative size of the solution matters, particularly
for the clique algorithm: if the solution is large (i.e.\ the two input graphs are very similar),
the clique approach finds nearly every labelled instance trivial.

@@

In many applications, the common subgraph must be connected. As illustrated in
\cref{figure:mcsexample}, the MCCS cannot be deduced from the MCS: we need to ensure connectedness
during the search. In \cref{mccs-cp}, we show that in CP this may be done in two different ways that
may be combined, and we show in \cref{mccs-cp-eval} that the best results are obtained when
combining them. In \cref{mccs-clique}, we introduce a new way for ensuring connectedness in a
clique-based approach. Finally, we compare CP and our clique-based approach in \cref{mccs-eval}.

@@
